A Tutorial on Tailored Simulation-Based Power Analysis for Experimental Designs with Generalized Linear Mixed Models

### 2023-10-23 docs/manuscript.Rmd
made final changes for submission

author: 
  - name          : "Timo K. Koch"
    affiliation   : "1, 2"
    corresponding : yes    # Define only one corresponding author
    address       : "Torstrasse 25, 9000 St. Gallen, Switzerland"
    email         : "timo.koch@unisg.ch"
    equal_contrib: yes
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - "Conceptualization"
      - "Formal Analysis"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
  - name          : "Florian Pargent"
    affiliation   : "1"
    equal_contrib: yes
    role:
      - "Formal Analysis"
      - "Methodology"
      - "Visualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
  - name          : "Anne-Kathrin Kleine"
    affiliation   : "1"
    role:
      - "Writing - Review & Editing"
  - name          : "Eva Lermer"
    affiliation   : "1,3"
    role:
      - "Resources"
  - name          : "Susanne Gaube"
    affiliation   : "1,4"
    role:
      - "Writing - Review & Editing"
      - "Supervision"

@@

  When planning experimental research, determining an appropriate sample size and using suitable statistical models are crucial for robust and informative results. However, the recent replication crisis underlines the need for more rigorous statistical methodology and well-powered designs. Generalized linear mixed models (GLMMs) offer a flexible statistical framework to analyze experimental data with complex (e.g., dependent and hierarchical) data structures. Yet, analytic methods and software cannot be applied to conduct a priori power analyses for GLMMs, necessitating data simulation approaches. Based on a practical case study, the current tutorial equips researchers with a step-by-step guide and corresponding code for conducting tailored a priori power analyses to determine appropriate sample sizes with GLMMs. Finally, we give an outlook on the increasing importance of simulation-based power analysis in experimental research.

@@

wordcount         : "7600"

@@

#csl               : "acm-sigchi-proceedings.csl" #added this for chi submission, not needed for preprint

@@

# Discussion

<!-- short summary of key takeaways from this tutorial -->
Experimental research requires careful planning and consideration of statistical power to ensure robust and meaningful results. While heuristics and user-friendly software can be useful for simple designs and models, they often fall short when more complex and customized simulations with GLMMs are required. The present tutorial presents a specific case study with corresponding code of how to conduct a simulation-based power analysis for experimental designs with GLMMs.

>>>>

author: 
  - name          : "Timo K. Koch"
    affiliation   : "1, 2"
    corresponding : yes    # Define only one corresponding author
    address       : "Torstrasse 25, 9000 St. Gallen, Switzerland"
    email         : "timo.koch@unisg.ch"
    equal_contrib: yes
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - "Conceptualization"
      - "Formal Analysis"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
  - name          : "Florian Pargent"
    affiliation   : "1"
    equal_contrib: yes
    role:
      - "Formal Analysis"
      - "Methodology"
      - "Visualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
  - name          : "Anne-Kathrin Kleine"
    affiliation   : "1"
    role:
      - "Formal Analysis"      
      - "Writing - Review & Editing"
  - name          : "Eva Lermer"
    affiliation   : "1,3"
    role:
      - "Conceptualization"
      - "Resources"
      - "Writing - Review & Editing"
  - name          : "Susanne Gaube"
    affiliation   : "1,4"
    role:
      - "Conceptualization"
      - "Supervision"
      - "Writing - Review & Editing"

@@

  When planning experimental research, determining an appropriate sample size and using suitable statistical models are crucial for robust and informative results. However, the recent replication crisis underlines the need for more rigorous statistical methodology and well-powered designs. Generalized linear mixed models (GLMMs) offer a flexible statistical framework to analyze experimental data with complex (e.g., dependent and hierarchical) data structures. Yet, available methods and software for a priori power analyses for GLMMs are often limited to specific designs, while data simulation approaches offer more flexibility. Based on a practical case study, the current tutorial equips researchers with a step-by-step guide and corresponding code for conducting tailored a priori power analyses to determine appropriate sample sizes with GLMMs. Finally, we give an outlook on the increasing importance of simulation-based power analysis in experimental research.

@@

wordcount         : "7504"

@@

@@
